---
title: "CourseCode"
output: html_document
---

### Example and R Commands
* `diamond` dataset from `UsingR` package
 - diamond prices in Singapore Dollars, diamond weight in carats (standard measure of diamond mass, 0.2g)
* `lm(price ~ I(carat - mean(carat)), data=diamond)` = mean centered linear regression
 - ***Note**: arithmetic operations must be enclosed in `I()` to work *
* `predict(fitModel, newdata=data.frame(carat=c(0, 1, 2)))` = returns predicted outcome from the given model (linear in our case) at the provided points within the `newdata` data frame
 - if `newdata` is unspecified (argument omitted), then `predict` function will return predicted values for all values of the predictor (x variable, carat in this case)
  + ***Note**: `newdata` has to be a dataframe, and the values you would like to predict (x variable, `carat` in this case) has to be specified, or the system won't know what to do with the provided values *
* `summary(fitModel)` = prints detailed summary of linear model
* ***example***
```{r echo=FALSE, results='hide', message=FALSE}
library(UsingR)
```

```{r}
# standard linear regression for price vs carat
fit <- lm(price ~ carat, data = diamond)
# intercept and slope
coef(fit)
# mean-centered regression
fit2 <- lm(price ~ I(carat - mean(carat)), data = diamond)
# intercept and slope
coef(fit2)
# regression with more granular scale (1/10th carat)
fit3 <- lm(price ~ I(carat * 10), data = diamond)
# intercept and slope
coef(fit3)
# predictions for 3 values
newx <- c(0.16, 0.27, 0.34)
# manual calculations
coef(fit)[1] + coef(fit)[2] * newx
# prediction using the predict function
predict(fit, newdata = data.frame(carat = newx))
```

* **interpretation**
 - we expect `r round(coef(fit)[2], 2)` (SIN) dollar increase in price for ***every carat increase*** in mass of diamond
 - or `r round(coef(fit3)[2], 2)` (SIN) dollar increase in price for ***every 1/10 carat*** increase in mass of diamond

* **prediction**
 - for 0.16, 0.27, and 0.34 carats, we predict the prices to be `r round(as.numeric(predict(fit, newdata = data.frame(carat = newx))),2)` (SIN) dollars


```{r, fig.height=4,fig.width=4, fig.align = 'center'}
# plot the data points
plot(diamond$carat, diamond$price, xlab = "Mass (carats)", ylab = "Price (SIN $)",
     bg = "lightblue", col = "black", cex = 1.1, pch = 21,frame = FALSE)
# plot linear regression line
abline(fit, lwd = 2)
# plot predictions for every value of carat (in red)
points(diamond$carat, predict(fit), pch = 19, col = "red")
# add guidelines for predictions for 0.16, 0.27, and 0.34
lines(c(0.16, 0.16, 0.12), c(200, coef(fit)[1] + coef(fit)[2] * 0.16,
      coef(fit)[1] + coef(fit)[2] * 0.16))
lines(c(0.27, 0.27, 0.12), c(200, coef(fit)[1] + coef(fit)[2] * 0.27,
      coef(fit)[1] + coef(fit)[2] * 0.27))
lines(c(0.34, 0.34, 0.12), c(200, coef(fit)[1] + coef(fit)[2] * 0.34,
      coef(fit)[1] + coef(fit)[2] * 0.34))
# add text labels
text(newx+c(0.03, 0, 0), rep(250, 3), labels = newx, pos = 2)
```


