---
title: "quiz3"
author: "Cliff Weaver"
date: "October 11, 2015"
output: html_document
---

Question 1

Consider the mtcars data set. Fit a model with mpg as the outcome that includes number of cylinders as a factor variable and weight as confounder. Give the adjusted estimate for the expected change in mpg comparing 8 cylinders to 4.

Solution:
```{r}
data(mtcars)
fit <- lm(mpg ~ factor(cyl) + wt, data = mtcars)
summary(fit)$coefficient
```
**ANSWER:  -6.071**

Question 2

Consider the mtcars data set. Fit a model with mpg as the outcome that includes number of cylinders as a factor variable and weight as a possible confounding variable. Compare the effect of 8 versus 4 cylinders on mpg for the adjusted and unadjusted by weight models. Here, adjusted means including the weight variable as a term in the regression model and unadjusted means the model without weight included. What can be said about the effect comparing 8 and 4 cylinders after looking at models with and without weight included?.

Solution:
```{r}
fit1 <- lm(mpg ~ as.factor(cyl), data = mtcars)
summary(fit1)$coef[3] 
## [1] -11.56364
summary(fit)$coef[3] 
## [1] -6.07086
```

**ANSWER:  Note that 11.564 > 6.071, and so holding weight constant, cylinder appears to have less of an impact on mpg than if weight is disregarded.**

Question 3

Consider the mtcars data set. Fit a model with mpg as the outcome that considers number of cylinders as a factor variable and weight as confounder. Now fit a second model with mpg as the outcome model that considers the interaction between number of cylinders (as a factor variable) and weight. Give the P-value for the likelihood ratio test comparing the two models and suggest a model using 0.05 as a type I error rate significance benchmark.

Solution:
```{r}
fit_inter <- lm(mpg ~ factor(cyl) * wt, data = mtcars)
anova(fit, fit_inter, test = "Chisq")
```
**ANSWER:  The P-value is larger than 0.05. So, according to our criterion, we would fail to reject, which suggests that the interaction terms may not be necessary.**


Question 4

Consider the mtcars data set. Fit a model with mpg as the outcome that includes number of cylinders as a factor variable and weight inlcuded in the model as

lm(mpg ~ I(wt * 0.5) + factor(cyl), data = mtcars)

How is the wt coefficient interpretted?

Solution:

**ANSWER:  Since the unit of (wt * 0.5) is (lb/2000), and one (short) ton is 2000 lbs, the wt coefficient is interpretted as the estimated expected change in MPG per one ton increase in weight for a specific number of cylinders (4, 6, 8).**
```{r}
fit4 <- lm(mpg ~ I(wt * 0.5) + factor(cyl), data = mtcars)
summary(fit4)$coefficient
```

Question 5

Consider the following data set

x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)

Give the hat diagonal for the most influential point.

Solution:
```{r}
x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)
fit5 <- lm(y ~ x)
hatvalues(fit5)
##         1         2         3         4         5 
## 0.2286650 0.2438146 0.2525027 0.2804443 0.9945734
```
**ANSWER:  0.9946**
Question 6

Consider the following data set

x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)

Give the slope dfbeta for the point with the highest hat value.

Solution:
```{r}
x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)
fit6 <- lm(y ~ x)
dfbetas(fit6)[, 2]
##             1             2             3             4             5 
##   -0.37811633   -0.02861769    0.00791512    0.67253246 -133.82261293
```
**ANSWER:  -134**  
Question 7

Consider a regression relationship between Y and X with and without adjustment for a third variable Z. Which of the following is true about comparing the regression coefficient between Y and X with and without adjustment for Z.

Solution:

**ANSWER:  It is possible for the coefficient to reverse sign after adjustment. For example, it can be strongly significant and positive before adjustment and strongly significant and negative after adjustment.**

