---
title: "R 4 Business"
output:
  html_document:
    highlight: pygments
    theme: spacelab
    toc: yes
---

<style type="text/css">
p{ /* Normal  */
   font-size: 14px;
   line-height: 18px;}
body{ /* Normal  */
   font-size: 14px;}
td {  /* Table  */
   font-size: 12px;}
h1 { /* Header 1 */
 font-size: 26px;
 color: #4294ce;}
h2 { /* Header 2 */
 font-size: 22px;}
h3 { /* Header 3 */
 font-size: 18px;}
code.r{ /* Code block */
  font-size: 12px;}
pre { /* Code block */
  font-size: 12px}
#table-of-contents h2 {
background-color: #4294ce;}
#table-of-contents{
background: #688FAD;}
#nav-top span.glyphicon{
color: #4294ce;}
#postamble{
background: #4294ce;
border-top: ;}
</style>

```{r setup, include=FALSE}
knitr::opts_knit$set(
     root.dir = 'C:/Users/czwea/Documents/GitHub/CourseWork/BusSciUniv/RforBusiness')
knitr::opts_chunk$set(eval=FALSE)
```
```{r echo=FALSE, warning=FALSE, message=FALSE}
#remotes::install_github("rstudio/gt")
rm(list = ls())
if(!require(easypackages)){install.packages("easypackages")}
library(easypackages)
packages("tidyverse", "gt", "here", "stringr", prompt = TRUE)
setwd("~/GitHub/CourseWork/BusSciUniv/RforBusiness")
```

# Business Case

2 objectives for a bicycle shop:

1. New Prodcut Idea
2. Segementation and Targeted Adverstisements

