#Week 4 Quiz

For this quiz we will be using several R packages. R package versions change over time, the right answers have been checked using the following versions of the packages. 

* AppliedPredictiveModeling: v1.1.6 
* caret: v6.0.47 
* ElemStatLearn: v2012.04-0 
* pgmm: v1.1 
* rpart: v4.1.8 
* gbm: v2.1 
* lubridate: v1.3.3 
* forecast: v5.6 
* e1071: v1.6.4 

If you aren't using these versions of the packages, your answers may not exactly match the right answer, but hopefully should be close. 

##Question 1

Load the vowel.train and vowel.test data sets:
```{r}
rm(list=ls()) #let's start doing this
library(ElemStatLearn)
library(caret)
library(gbm)
data(vowel.train)
data(vowel.test) 
```
Set the variable y to be a factor variable in both the training and test set. Then set the seed to 33833. Fit (1) a random forest predictor relating the factor variable y to the remaining variables and (2) a boosted predictor using the "gbm" method. Fit these both with the train() command in the caret package. 
What are the accuracies for the two approaches on the test data set? What is the accuracy among the test set samples where the two methods agree? 

###Solution 1  
Not sure this one works
```{r}
vowel.train$y <- factor(vowel.train$y)
vowel.test$y <- factor(vowel.test$y)

set.seed(33833)

# create models
fit1 <- train(y ~ ., data = vowel.train, method = "rf", trControl = trainControl(number = 4))
fit2 <- train(y ~ ., data = vowel.train, method = "gbm", verbose=FALSE) #without Verbose lots of data scrolls

# predict test
predict1 <- predict(fit1, newdata = vowel.test)
predict2 <- predict(fit2, newdata = vowel.test)

# combine predictions
DF_combined <- data.frame(predict1, predict2, y = vowel.test$y)
fit_combined <- train(y ~ ., data = DF_combined, method = "gam")
predict3 <- predict(fit_combined, newdata = vowel.test)

# confusion matrixes
c1 <- confusionMatrix(predict1, vowel.test$y) #Accuracy : 0.6017 
c2 <- confusionMatrix(predict2, vowel.test$y) #Accuracy : 0.5173
c3 <- confusionMatrix(predict3, DF_combined$y) #Accuracy : 0.1537
```
###Solution 2  
See https://rpubs.com/calin/predmachlearn-033-finalQuiz
```{r}
#vowel.train_2$y <- as.factor(vowel.train$y) Already performed in the 1st solution
#vowel.test_2$y <- as.factor(vowel.test$y)
set.seed(33833)
# fit a random forest predictor using caret's unified model building interface
vowel.fit.rf <- train(y ~ .,
                      data = vowel.train,
                      method="rf")

# fit a gradient boosting machine model, using verbose to limit the output
vowel.fit.gbm <- train(y ~ .,
                      data = vowel.train,
                      method="gbm", 
                      verbose = F)
```
We’ll look at the Out of Sample error for both random forest and grandient boosting machine models.
```{r}
# Random Forest: 100% accuracy on the training set
confusionMatrix(predict(vowel.fit.rf), vowel.train$y) #Accuracy : 1
```
```{r}
# but only 60.61% accuracy on the testing set
confusionMatrix(predict(vowel.fit.rf, vowel.test), vowel.test$y) #Accuracy : 0.6039
```
```{r}
# GBM now: 100% accuracy on training
confusionMatrix(predict(vowel.fit.gbm), vowel.train$y) #Accuracy : 1
```
```{r}
# Accuracy: 0.526
confusionMatrix(predict(vowel.fit.gbm, vowel.test), vowel.test$y) #Accuracy : 0.5108
```
therefore both are largely overfitted model.

##Question 2
(Could look at a related problem here:  http://rseiter.github.io/PracticalMLProject/CaretEnsembleExample.html)
Load the Alzheimer's data using the following commands library(caret) library(gbm)
```{r}
rm(list=ls()) #let's start doing this - still need to run the code provided with the problem
set.seed(3433)
library(AppliedPredictiveModeling)
data(AlzheimerDisease)
adData = data.frame(diagnosis,predictors)
inTrain = createDataPartition(adData$diagnosis, p = 3/4)[[1]]
training = adData[ inTrain,]
testing = adData[-inTrain,]
```
Set the seed to 62433 and predict diagnosis with all the other variables using a random forest ("rf"), boosted trees ("gbm") and linear discriminant analysis ("lda") model. Stack the predictions together using random forests ("rf"). What is the resulting accuracy on the test set? Is it better or worse than each of the individual predictions? 

###Solution 1
```{r}
dim(adData) # 333 131
# head(adData)
set.seed(62433)
fitRf <- train(diagnosis ~ ., data=training, method="rf")
fitGBM <- train(diagnosis ~ ., data=training, method="gbm", verbose = FALSE)
fitLDA <- train(diagnosis ~ ., data=training, method="lda")
predRf <- predict(fitRf, testing)
predGBM <- predict(fitGBM, testing)
predLDA <- predict(fitLDA, testing)
pred <- data.frame(predRf, predGBM, predLDA, diagnosis=testing$diagnosis)
# Stack the predictions together using random forests ("rf")
fit <- train(diagnosis ~., data=pred, method="rf")
predFit <- predict(fit, testing)
c1_1 <- confusionMatrix(predRf, testing$diagnosis)$overall[1]
c2_1 <- confusionMatrix(predGBM, testing$diagnosis)$overall[1]
c3_1 <- confusionMatrix(predLDA, testing$diagnosis)$overall[1]
c4_1 <- confusionMatrix(predFit, testing$diagnosis)$overall[1]
print(paste(c1_1, c2_1, c3_1, c4_1)) 
```
Stacked Accuracy: 0.79 is better than random forests and lda and the same as boosting.
```
###Solution 2
```{r}
set.seed(62433)

# create models
fit1 <- train(diagnosis ~ ., data = training, method = "rf", trControl = trainControl(number = 4))
fit2 <- train(diagnosis ~ ., data = training, method = "gbm", verbose=FALSE)
fit3 <- train(diagnosis ~ ., data = training, method = "lda")

# predict test
predict1 <- predict(fit1, newdata = testing)
predict2 <- predict(fit2, newdata = testing)
predict3 <- predict(fit3, newdata = testing)

# combine predictions
DF_combined <- data.frame(predict1, predict2, predict3, diagnosis = testing$diagnosis) # training$diagnosis?
fit_combined <- train(diagnosis ~ ., data = DF_combined, method = "rf")
predict4 <- predict(fit_combined, newdata = testing)

# confusion matrixes
c1_2 <- confusionMatrix(predict1, testing$diagnosis)
c2_2 <- confusionMatrix(predict2, testing$diagnosis)
c3_2 <- confusionMatrix(predict3, testing$diagnosis)
c4_2 <- confusionMatrix(predict4, testing$diagnosis)

print(paste(c1_2$overall[1], c2_2$overall[1], c3_2$overall[1], c4_2$overall[1]))
```
###Solution 3
```{r}
# rough data estimates
dim(training); dim(testing);
```
```{r}
set.seed(62433)
# fit a Random Forest model
alz.fit.rf = train(diagnosis ~ .,
                   data = training,
                   method = "rf")

# fit a Gradient Boosting Machine model
alz.fit.gbm = train(diagnosis ~ .,
                   data = training,
                   method = "gbm", verbose = FALSE)

# fit a Linear Discriminant Analysis model
alz.fit.lda = train(diagnosis ~ .,
                   data = training,
                   method = "lda")
# now we evaluate the accuracy for all 3 models individualy

# Random Forest
postResample(predict(alz.fit.rf, testing), testing$diagnosis)
```
```{r}
# Gradient Boosting Machine
postResample(predict(alz.fit.gbm, testing), testing$diagnosis)
```
```{r}
# Linear Discriminant Analysis
postResample(predict(alz.fit.lda, testing), testing$diagnosis)
```
Stacking the 3 models using Random Forest  
we fit a model that combines predictors
```{r}
# build a data frame with the combined predictions and the outcome variable
predDF = data.frame(
                  predict(alz.fit.rf, testing),
                  predict(alz.fit.gbm, testing),
                  predict(alz.fit.lda, testing),
                  diagnosis = testing$diagnosis)
dim(predDF); head(predDF)
```
```{r}
# training a stacking model using random forest
combModFit <- train(diagnosis ~ ., method = "rf", data=predDF)
```
```{r}
# predicting on the combined data frame
combPred <- predict(combModFit, predDF)
# resolve overall agreement rate and Kappa
postResample(combPred, testing$diagnosis)
```
##Question 3
Load the concrete data with the commands: 
```{r}
rm(list=ls()) #let's start doing this
set.seed(3523)
library(AppliedPredictiveModeling)
data(concrete)
inTrain = createDataPartition(concrete$CompressiveStrength, p = 3/4)[[1]]
training = concrete[ inTrain,]
testing = concrete[-inTrain,]
```
Set the seed to 233 and fit a lasso model to predict Compressive Strength. Which variable is the last coefficient to be set to zero as the penalty increases? (Hint: it may be useful to look up ?plot.enet). 

###Solution 1
```{r}
set.seed(233)
fit_1 <- train(CompressiveStrength ~ ., data=training, method="lasso")
fit_1
plot.enet(fit_1$finalModel, xvar="penalty", use.color=T) # Cement
```
###Solution 2
```{r}
set.seed(233)

fit_2 <- train(CompressiveStrength ~ ., data = training, method = "lasso")

# Since we are interested in the shrinkage of coefficients as the penalty(lambda) increases, "
# penalty" looks promising for an xvar argument value.
plot.enet(fit_2$finalModel, xvar = "penalty", use.color = TRUE)
```
###Solution 3
```{r}
set.seed(233)
conc.fit.lasso = train(CompressiveStrength ~ .,
                       data = training,
                       method = "lasso")
plot.enet(conc.fit.lasso$finalModel,
          xvar="penalty", use.color=TRUE)
```
```{r}
# the chart doesn't say much, however this table reveals the last "zeroed" coefficient
conc.fit.lasso$finalModel$beta.pure
```
##Question 4

Load the data on the number of visitors to the instructors blog from here: 
https://d396qusza40orc.cloudfront.net/predmachlearn/gaData.csv

Using the commands:
```{r}
rm(list=ls()) #let's start doing this
library(lubridate)  # For year() function below
dat = read.csv("./data/gaData.csv")
training = dat[year(dat$date) < 2012,]
testing = dat[(year(dat$date)) > 2011,]
tstrain = ts(training$visitsTumblr)
```
Fit a model using the bats() function in the forecast package to the training time series. Then forecast this model for the remaining time points. For how many of the testing points is the true value within the 95% prediction interval bounds? 
###Solution 1:
```{r}
library(forecast)
library(quantmod)
fit_1 <- bats(tstrain)
fit_1
pred <- forecast(fit_1, level=95, h=dim(testing)[1])
names(data.frame(pred))
predComb <- cbind(testing, data.frame(pred))
names(testing)
names(predComb)
predComb$in95 <- (predComb$Lo.95 < predComb$visitsTumblr) & 
                    (predComb$visitsTumblr < predComb$Hi.95)
# How many of the testing points is the true value within the 
# 95% prediction interval bounds?
prop.table(table(predComb$in95))[2] # 0.9617021

```
###Solution 2
```{r}
library(forecast)
library(quantmod)

# fit a model
fit_2 <- bats(tstrain)

# check how long the test set is, so you can predict beyond trainign
h <- dim(testing)[1]

# forecast the model for remaining time points
fcast <- forecast(fit, level = 95, h = h)

# get the accuracy
accuracy(fcast, testing$visitsTumblr)


# check what percentage of times that the actual number of visitors was within
# 95% confidence interval

result <- c()
l <- length(fcast$lower)

for (i in 1:l){
  x <- testing$visitsTumblr[i]
  a <- fcast$lower[i] < x & x < fcast$upper[i]
  result <- c(result, a)
}

sum(result)/l * 100
```
###Solution 3

```{r}
library(forecast)

# build a bats model based on the original time series
visits.exp.smoothing = bats(tstrain)
# build the forecast with the same range as the testing set (2012)
visits.forecast = forecast(visits.exp.smoothing, nrow(testing))
# plot the forecast
plot(visits.forecast)
```

```{r}
# extracting the 95% prediction boundaries
visits.forecast.lower95 = visits.forecast$lower[,2]
visits.forecast.upper95 = visits.forecast$upper[,2]

# see how many of the testing visit counts do actually match
table ( 
  (testing$visitsTumblr>visits.forecast.lower95) & 
  (testing$visitsTumblr<visits.forecast.upper95))
```
```{r}
# and in percentages
226/nrow(testing)
```

##Question 5

Load the concrete data with the commands: 
```{r}
rm(list=ls()) #let's start doing this - still need to run the code provided with the problem
set.seed(3523)
library(AppliedPredictiveModeling)
data(concrete)
inTrain = createDataPartition(concrete$CompressiveStrength, p = 3/4)[[1]]
training = concrete[ inTrain,]
testing = concrete[-inTrain,]
```
Set the seed to 325 and fit a support vector machine using the e1071 package to predict Compressive Strength using the default settings. Predict on the testing set. What is the RMSE? 
###Solution 1
```{r}
set.seed(325)
fit <- svm(CompressiveStrength ~., data=training)
# OR another way
# fit_1 <- train(CompressiveStrength ~. data=training, method="svmRadial")
pred <- predict(fit_1, testing)
acc <- accuracy(pred, testing$CompressiveStrength)
acc
acc[2] # RMSE 6.715009
```
###Solution 2
```{r}
set.seed(325)

library(e1071)
library(caret)

fit_2 <- train(CompressiveStrength ~ ., data = training, method = "svmRadial")

prediction <- predict(fit_2, testing)

accuracy(prediction, testing$CompressiveStrength)
```
###Solution 3
```{r}
#install.packages("e1071")
library("e1071")
set.seed(325)
conc.fit.svm = svm(CompressiveStrength ~ .,
                     data=training)
# comparing predictions to actual values
conc.pred.svm = predict(conc.fit.svm, newdata = testing)

# Root Mean Squared Error
error = conc.pred.svm - testing$CompressiveStrength
sqrt(mean(error^2))
```
```{r}
# plot the relationship between the forecasted svm values and the actual values, coloured by Age
plot(conc.pred.svm, testing$CompressiveStrength, 
              pch=20, cex=1, 
              col=testing$Age,
              main="Relationship between the svm forecast and actual values")
```
