---
title: "Data manipulation with tidyr"
output:
  html_document:
    highlight: pygments
    theme: spacelab
    toc: yes
header-includes: \usepackage{graphicx} \usepackage{mathtools}
---
```{r echo=FALSE, message=FALSE}
if(!require(easypackages)){
    install.packages("easypackages")
    library(easypackages)
}
packages("dplyr", "tidyr", prompt = FALSE)
```

## Introduction

`tidyr` is a package by Hadley Wickham that makes it easy to tidy your data. It is often used in conjunction with `dplyr`. Data is tidy when each column represents a variable and each row represents an observation.

`tidyr` is part of the `tidyverse`.  `tidyverse` is a set of packages that work in harmony because they share common data representations and API design. `tidyverse` is designed to make it easy to [install and load](https://blog.rstudio.org/2016/09/15/tidyverse-1-0-0/) core packages from the `tidyverse` in a single command.  I recommend you become intimately familiar with these tools.

Explore `tidyr`functions:

* gather – converts wide data to longer format. It is analogous to `melt` in reshape2.
* spread – converts long data to wider format. It is analogous to the `cast` in reshape2.
* unite – combines two or more columns into a single column.
* separate – splits one column into two or more columns.

> reshape2 is another popular R package.  Like many function in R, there are many ways to accomplish any given task.  With all due respect to reshape2, I prefer to use the tools in `tidyverse`.

## Get Data

Use the mtcars dataset:
```{r message=FALSE, warning=FALSE}
head(mtcars)
```

Include the names of the cars in a column called car.

```{r}
mtcars$car <- rownames(mtcars)
mtcars <- mtcars[, c(12, 1:11)]
head(mtcars)
```

## tidyr Functions
### tidyr - gather

`gather` takes the form: `gather(data, key, value, ..., na.rm = FALSE, convert = FALSE)`

where ... is the specification of the columns to gather.

```{r}
mtcarsNew <- mtcars %>% gather(attribute, value, -car)
head(mtcarsNew)
```

It gathers all the columns except car and places their name and value into the attritube and value column.

The great thing about `tidyr` is that you can gather only certain columns and leave the others alone. If we want to gather all the columns from mpg to gear and leave the carb and car columns as they are we do this:

```{r}
mtcarsNew <- mtcars %>% gather(attribute, value, mpg:gear)
head(mtcarsNew)
```

### tidyr - spread

`spread` takes the form: `spread(data, key, value, fill = NA, convert = FALSE, drop = TRUE)`

Here is an example:

```{r}
mtcarsSpread <- mtcarsNew %>% spread(attribute, value)
head(mtcarsSpread)
```

### tidyr - unite

`unite` takes the form: `unite(data, col, ..., sep = "_", remove = TRUE)`

where ... represents the columns to unite and col represents the column to add.

We need new data to demo this::
```{r}
date <- as.Date('2017-07-01') + 0:14
hour <- sample(1:24, 15)
min <- sample(1:60, 15)
second <- sample(1:60, 15)
event <- sample(letters, 15)
data <- data.frame(date, hour, min, second, event)
head(data,5)
```

Combine the date, hour, min, and second columns into a new column called datetime. Datetime in R is of the form Year-Month-Day Hour:Min:Second.

```{r}
dataNew <- data %>% unite(datehour, date, hour, sep = ' ') %>% unite(datetime, datehour, min, second, sep = ':')
head(dataNew, 5)
```

### tidyr - separate

`separate` takes the form: `separate(data, col, into, sep = "[^[:alnum:]]+", remove = TRUE, convert = FALSE, extra = "warn", fill = "warn", ...)`

Revert back to the original data using `separate`:

```{r}
data1 <- dataNew %>% separate(datetime, c('date', 'time'), sep = ' ') %>% separate(time, c('hour', 'min', 'second'), sep = ':')
head(data1,5)
```

It first splits the datetime column into date and time, and then splits time into hour, min, and second.
