---
title: "Hierarchial Clustering French Cities"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

See http://www.r-bloggers.com/clustering-french-cities-based-on-temperatures/?utm_source=feedburner&utm_medium=email&utm_campaign=Feed%3A+RBloggers+%28R+bloggers%29

In order to illustrate hierarchical clustering techniques and k-means, I did borrow François Husson‘s dataset, with monthly average temperature in several French cities.
```{r}
temp=read.table("http://freakonometrics.free.fr/FR_temp.txt", header=TRUE,dec=",")
```
We have 15 cities, with monthly observations
```{r}
X=temp[,1:12]
boxplot(X)
```

Since the variance seems to be rather stable, we will not ‘normalize’ the variables here,
```{r}
apply(X,2,sd)
```
In order to get a hierarchical cluster analysis, use for instance
```{r message=FALSE}
h <- hclust(dist(X), method = "ward")
plot(h, labels = rownames(X), sub = "")
```

An alternative is to use
```{r message=FALSE}
library(FactoMineR)
h2=HCPC(X)
dev.off() #needed to add this to avoid "Error in plot.new() : figure margins too large"
plot(h2)
```
Here, we visualise observations with a principal components analysis. We have here also an automatic selection of the number of classes, here 3. We can get the description of the groups using
```{r}
h2$desc.ind
```

or directly
```{r}
cah=hclust(dist(X))
groups.3 <- cutree(cah,3)
```

We can also visualise those classes by ourselves,
```{r}
acp=PCA(X,scale.unit=FALSE)
plot(acp$ind$coord[,1:2],col="white")
text(acp$ind$coord[,1],acp$ind$coord[,2], rownames(acp$ind$coord),col=groups.3)

#It is possible to plot the centroïds of those clusters

PT=aggregate(acp$ind$coord,list(groups.3),mean)
points(PT$Dim.1,PT$Dim.2,pch=19)
```

If we add Voroid sets around those centroïds, here we do not see them (actually, we see the point – in the middle – that is exactly at the intersection of the three regions),
```{r}
library(tripack)
plot(acp$ind$coord[,1:2],col="white")
text(acp$ind$coord[,1],acp$ind$coord[,2], rownames(acp$ind$coord),col=groups.3)
V <- voronoi.mosaic(PT$Dim.1,PT$Dim.2)
plot(V,add=TRUE)
```

To visualize those regions, use
```{r}
p=function(x,y){which.min((PT$Dim.1-x)^2+(PT$Dim.2-y)^2)}

vx=seq(-10,12,length=251)
vy=seq(-6,8,length=251)
z=outer(vx,vy,Vectorize(p))
image(vx,vy,z,col=c(rgb(1,0,0,.2), rgb(0,1,0,.2),rgb(0,0,1,.2)))
CL=c("red","black","blue")
text(acp$ind$coord[,1],acp$ind$coord[,2], rownames(acp$ind$coord),col=CL[groups.3])
```

Actually, those three groups (and those three regions) are also the ones we obtain using a k-mean algorithm,
```{r}
km=kmeans(acp$ind$coord[,1:2],3)
km
```

But actually, since again we have some spatial data, it is possible to visualize them on a map
```{r warning=FALSE, message=FALSE}
library(maps)
map("france")
points(temp$Long,temp$Lati,col=groups.3,pch=19)
```

or, to visualize the regions, use e.g.
```{r warning=FALSE}
library(car)
map("france")
points(temp$Long,temp$Lati,col=groups.3,pch=19)
for(i in 1:3) 
dataEllipse(temp$Long[groups.3==i], temp$Lati[groups.3==i], levels=.7,add=TRUE, col=i+1,fill=TRUE)
