---
title: "ISLR - Ch 6 - Linear Model Selection and Regularization"
output: html_document
---

##Best Subset Selection

Apply the best subset selection approach to the Hitters data. Predict a baseball player’s Salary on the basis of various statistics associated with performance in the previous year.

The Salary variable is missing for some of the players. The is.na() function can be used to identify the missing observations. It returns a vector of the same length as the input vector, with a TRUE for any elements that are missing, and a FALSE for non-missing elements. The sum() function can then be used to count all of the missing elements.
```{r}
library(ISLR)
head(Hitters)
names(Hitters)
dim(Hitters)
sum(is.na(Hitters$Salary))
```
Salary is missing for 59 players. The na.omit() function removes all of the rows that have missing values in any variable.
```{r}
Hitters = na.omit(Hitters)
dim(Hitters)
sum(is.na(Hitters))
```
The **regsubsets()** function (part of the **leaps library**) performs best set selection by identifying the best model that contains a given number of predictors, where best is quantified using RSS. The syntax is the same as for lm(). The summary() command outputs the best set of variables for each model size.
```{r}
library(leaps)
regfit.full = regsubsets(Salary ~., Hitters)
summary(regfit.full)
```
An asterisk indicates that a given variable is included in the corresponding model. For instance, this output indicates that the best two-variable model contains only Hits and CRBI. By default, regsubsets() only reports results up to the best eight-variable model. But the nvmax option can be used in order to return as many variables as are desired. Fit up to a 19-variable model.
```{r}
regfit.full = regsubsets(Salary ~., data = Hitters, nvmax = 19)
reg.summary = summary(regfit.full)
```
The summary() function also returns R2, RSS, adjusted R2, Cp, and BIC. We can examine these to try to select the best overall model.
```{r}
names(reg.summary )
```
Rhe R2 statistic increases from 32% when one variable is included in the model, to almost 55 % when all variables are included. The R2 increases monotonically as more variables are included.
```{r}
reg.summary$rsq
```
Plotting RSS, adjusted R2, Cp, and BIC for all of the models at once will help determine which model to select. type="l" option tells R to connect the plotted points with lines.
```{r}
par(mfrow = c(2, 2))
plot(reg.summary$rss, xlab = "Number of Variables", ylab ="RSS", type="l")
plot(reg.summary$adjr2, xlab = "Number of Variables", ylab ="Adjusted RSq", type="l")
which.max(reg.summary$adjr2)
points(11, reg.summary$adjr2[11], col = "red", cex = 2, pch = 20)
```

The points() command works like the plot() command, except that it points() puts points on a plot that has already been created, instead of creating a new plot. The which.max() function can be used to identify the location of the maximum point of a vector. We will now plot a red dot to indicate the model with the largest adjusted R2 statistic.

In a similar fashion we can plot the Cp and BIC statistics and indicate the models with the smallest statistic using which.min().
```{r}
plot(reg.summary$cp, xlab = "Number of Variables", ylab="Cp", type="l")
which.min(reg.summary$cp)

points(10, reg.summary$cp[10], col = "red", cex = 2, pch = 20)
which.min(reg.summary$bic)

plot(reg.summary$bic, xlab = "Number of Variables", ylab = "BIC", type="l")
points(6, reg.summary$bic[6], col = "red", cex = 2, pch =20 )
```

The regsubsets() function has a built-in plot() command which can be used to display the selected variables for the best model with a given number of predictors, ranked according to the BIC, Cp, adjusted R2, or AIC. To find out more about this function, type ?plot.regsubsets. 
```{r}
plot(regfit.full, scale = "r2")
plot(regfit.full, scale = "adjr2")
plot(regfit.full, scale = "Cp")
plot(regfit.full, scale = "bic")
```

The top row of each plot contains a black square for each variable selected according to the optimal model associated with that statistic. Sseveral models share a BIC close to −150. However, the model with the lowest BIC is the six-variable model that contains only AtBat, Hits, Walks, CRBI, DivisionW, and PutOuts. Use the coef() function to see the coefficient estimates associated with this model.
```{r}
coef(regfit.full, 6)
```
###Forward and Backward Stepwise Selection

Use the regsubsets() function to perform forward stepwise or backward stepwise selection, using the argument method="forward" or method="backward".
```{r}
regfit.fwd = regsubsets(Salary ~., data = Hitters, nvmax = 19, method = "forward")
summary(regfit.fwd)

regfit.bwd = regsubsets(Salary ~., data = Hitters, nvmax = 19, method = "backward")
summary(regfit.bwd )
```
Using forward stepwise selection, the best onevariable model contains only CRBI, and the best two-variable model additionally includes Hits. For this data, the best one-variable through sixvariable models are each identical for best subset and forward selection. However, the best seven-variable models identified by forward stepwise selection, backward stepwise selection, and best subset selection are different.
```{r}
coef(regfit.full, 7)
coef(regfit.fwd, 7)
coef(regfit.bwd, 7)
```
###Choosing Among Models Using the Validation Set Approach and Cross-Validation

It is possible to choose among a set of models of different sizes using Cp, BIC, and adjusted R2. Using the validation set and cross-validation approaches to yield accurate estimates of the test error, use only the training observations to perform all aspects of model-fitting—including variable selection. Therefore, the determination of which model of a given size is best must be made using only the training observations. This point is subtle but important. If the full data set is used to perform the best subset selection step, the validation set errors and cross-validation errors that we obtain will not be accurate estimates of the test error. 

In order to use the validation set approach, begin by splitting the observations into a training set and a test set. Create a random vector, train, of elements equal to TRUE if the corresponding observation is in the training set, and FALSE otherwise. The vector test has a TRUE if the observation is in the test set, and a FALSE otherwise. Note the ! in the command to create test causes TRUEs to be switched to FALSEs and vice versa. Set a random seed so that the user will obtain the same training set/test set split.
```{r}
set.seed(1)
train = sample(c(TRUE, FALSE), nrow(Hitters), rep = TRUE)
test = (!train)
```

Apply regsubsets() to the training set in order to perform best subset selection.
```{r}
regfit.best = regsubsets(Salary ~., data = Hitters[train, ], nvmax = 19)
```

Compute the validation set error for the best model of each model size. We first make a model matrix from the test data.
```{r}
test.mat = model.matrix(Salary ~.,data = Hitters[test, ])
```
The **model.matrix()** function is used in many regression packages for building an “X” matrix from data. Run a loop and for each size i, extract the coefficients from regfit.best for the best model of that size, multiply them into the appropriate columns of the test model matrix to form the predictions, and compute the test MSE.
```{r}
val.errors = rep(NA, 19)
for(i in 1:19){coefi=coef(regfit.best, id = i)
     pred = test.mat[, names(coefi)]%*% coefi
     val.errors [i]= mean(( Hitters$Salary[test]-pred)^2)
}
```
The best model is the one that contains ten variables.
```{r}
val.errors
which.min(val.errors)
coef(regfit.best, 10)
```
This is little tedious, partly because there is no predict() method for regsubsets(). Capture the steps above and write a predict method.
```{r}
predict.regsubsets =function(object ,newdata ,id ,...){
   form=as.formula (object$call [[2]])
   mat=model.matrix (form ,newdata )
   coefi =coef(object ,id=id)
   xvars =names (coefi )
   mat[,xvars ]%*% coefi
}
```
The  complex part is how the formula used in the call to regsubsets(). It demonstrate how to use this function below, when doing cross-validation. 

Perform best subset selection on the full data set and select the best ten-variable model. It is important that to make use of the full data set in order to obtain more accurate coefficient estimates. Note that best subset selection is used on the full data set and select the best tenvariable model rather than using the variables that were obtained from the training set, because the best ten-variable model on the full data set may differ from the corresponding model on the training set.
```{r}
regfit.best = regsubsets(Salary ~., data = Hitters, nvmax = 19)
coef(regfit.best, 10)
```

The best ten-variable model on the full data set has a diﬀerent set of variables than the best ten-variable model on the training set. 

Choose among the models of diﬀerent sizes using crossvalidation. This approach is somewhat involved. Must perform best subset selection within each of the k training sets. With its clever subsetting syntax, R makes this job quite easy. First, create a vector that allocates each observation to one of k = 10 folds and create a matrix to store the results.
```{r}
k=10
set.seed(1)
fold = sample(1:k, nrow(Hitters), replace=TRUE)
cv.errors = matrix(NA, k , 19, dimnames =list(NULL, paste(1:19)))
```

Write a for loop that performs cross-validation. In the jth fold the elements of folds that equal j are in the test set and the remainder are in the training set. Make predictions for each model size (using the new predict() method) compute the test errors on the appropriate subset and store them in the appropriate slot in the matrix cv.errors.
```{r}
for(j in 1:k){
     best.fit = regsubsets(Salary~., data=Hitters[fold!=j, ], 
          nvmax=19)
     for(i in 1:19){ 
          pred=predict(best.fit, Hitters[fold==j,],id=i)
          cv.errors[j,i]=mean((Hitters$Salary[fold==j]-pred)^2)
     }
}
```

This provides a 10×19 matrix, of which the (i,j)th element corresponds to the test MSE for the ith cross-validation fold for the best j-variable model. Use the apply() function to average over the columns of this matrix in order to obtain a vector for which the jth element is the crossvalidation error for the j-variable model.
```{r}
mean.cv.errors=apply(cv.errors, 2, mean)
mean.cv.errors
par(mfrow=c(1,1))
plot(mean.cv.errors, type="b")
```

The cross-validation selects an 11-variable model. Perform best subset selection on the full data set in order to obtain the 11-variable model.
```{r}
reg.best=regsubsets(Salary~., data=Hitters, nvmax=19)
coef(reg.best, 11) 
```

##Ridge Regression and the Lasso

Use the glmnet package in order to perform ridge regression and the lasso. The main function in this package is glmnet(), which can be used to ﬁt ridge regression models, lasso models, and more. This function has slightly diﬀerent syntax from other model-ﬁtting functions. In particular, pass in an x matrix as well as a y vector and  do not use the y ∼ x syntax.Nnow perform ridge regression and the lasso in order to predict Salary on the Hitters data. Before proceeding ensure that the missing values have been removed from the data.
```{r}
x = model.matrix(Salary~., Hitters)[, -1]
y=Hitters$Salary
```

The **model.matrix()** function is particularly useful for creating x; not only does it produce a matrix corresponding to the 19 predictors but it also automatically transforms any qualitative variables into dummy variables. The latter property is important because glmnet() can only take numerical, quantitative inputs.

###Ridge Regression

The glmnet() function has an alpha argument that determines what type of model is ﬁt. If alpha=0 then a ridge regression model is ﬁt and if alpha=1 then a lasso model is ﬁt. Frst ﬁt a ridge regression model.
```{r}
library(glmnet)
grid=10^seq(10,-2, length =100)
ridge.mod=glmnet(x,y,alpha=0,lambda=grid)
```

By default the glmnet() function performs ridge regression for an automatically selected range of λ values. To implement the function over a grid of values ranging from λ = 1010 to λ = 10 −2, essentially covering the full range of scenarios from the null model containing only the intercept to the least squares ﬁt. Also compute model ﬁts for a particular value of λ that is not one of the original grid values. By default, the glmnet() function standardizes the variables so that they are on the same scale. To turn oﬀ this default setting, use the argument standardize=FALSE. Associated with each value of λ is a vector of ridge regression coeﬃcients, stored in a matrix that can be accessed by coef(). In this case, it is a 20×100 matrix, with 20 rows (one for each predictor, plus an intercept) and 100 columns (one for each value of λ).
```{r}
dim(coef(ridge.mod))
```
The coeﬃcient estimates to be much smaller in terms of l~2~ norm, when a large value of λ is used, as compared to when a small value of λ is used. These are the coeﬃcients when λ = 11 ,498, along with their l~2~ norm:
```{r}
ridge.mod$lambda[50] 
coef(ridge.mod)[,50]
sqrt(sum(coef(ridge.mod)[-1,50]^2) )
```
In contrast, here are the coeﬃcients when λ = 705 along with their l~2~ norm. Note the much larger l~2~ norm of the coeﬃcients associated with this smaller value of λ.
```{r}
ridge.mod$lambda [60] 
coef(ridge.mod)[,60] 
sqrt(sum(coef(ridge.mod)[-1,60]^2))
```
Use the predict() function for a number of purposes. Obtain the ridge regression coeﬃcients for a new value of λ, say 50:
```{r}
predict(ridge.mod, s=50, type="coefficients")[1:20,] 
```
Split the samples into a training set and a test set in order to estimate the test error of ridge regression and the lasso. There are two common ways to randomly split a data set. The ﬁrst is to produce a random vector of TRUE, FALSE elements and select the observations corresponding to TRUE for the training data. The second is to randomly choose a subset of numbers between 1 and n; these can then be used as the indices for the training observations. The two approaches work equally well. Demonstrate the latter approach. Set a random seed so that the results obtained will be reproducible.
```{r}
set.seed(1)
train=sample(1:nrow(x), nrow(x)/2)
test=(-train)
y.test=y[test]
```

Fit a ridge regression model on the training set and evaluate its MSE on the test set, using λ = 4. Note the use of the predict() function again. Get predictions for a test set by replacing type="coefficients" with the newx argument.
```{r}
ridge.mod=glmnet(x[train,],y[train],alpha=0,lambda=grid, thresh=1e-12)
ridge.pred=predict(ridge.mod, s=4, newx=x[test ,])
mean((ridge.pred -y.test)^2) 
```
The test MSE is 101037. If had instead simply ﬁt a model with just an intercept would have predicted each test observation using the mean of the training observations. In that case, compute the test set MSE like this:
```{r}
mean((mean(y[train])-y.test)^2)
```
Could also get the same result by ﬁtting a ridge regression model with a very large value of λ. Note that 1e10 means 10~10~.
```{r}
ridge.pred=predict(ridge.mod, s=1e10, newx=x[test,])
mean((ridge.pred-y.test)^2)
```
So ﬁtting a ridge regression model with λ = 4 leads to a much lower test MSE than ﬁtting a model with just an intercept. Now check whether there is any beneﬁt to performing ridge regression with λ = 4 instead of just performing least squares regression. Recall that least squares is simply ridge regression with λ = 0.
```{r}
ridge.pred=predict(ridge.mod, s=0,newx=x[test ,],exact=T)
mean((ridge.pred-y.test)^2)
lm(y~x, subset=train)
predict(ridge.mod, s=0, exact=TRUE, type="coefficients")[1:20,]
```
To ﬁt a (unpenalized) least squares model, then use the lm() function since that function provides more useful outputs such as standard errors and p-values for the coeﬃcients. 

In general, instead of arbitrarily choosing λ = 4, it would be better to use cross-validation to choose the tuning parameter λ. Do this using the built-in cross-validation function, cv.glmnet(). By default, the function performs ten-fold cross-validation, though this can be changed using the argument nfolds. Set a random seed ﬁrst so the results will be reproducible since the choice of the cross-validation folds is random.
```{r}
set.seed(1)
cv.out=cv.glmnet(x[train,],y[train],alpha=0)
plot(cv.out)
bestlam =cv.out$lambda.min
bestlam
```
The value of λ that results in the smallest crossvalidation error is 212. What is the test MSE associated with this value of λ?
```{r}
ridge.pred=predict(ridge.mod ,s=bestlam ,newx=x[test,])
mean((ridge.pred-y.test)^2)
```
This represents a further improvement over the test MSE that we got using λ = 4. Finally,  reﬁt our ridge regression model on the full data set using the value of λ chosen by cross-validation and examine the coeﬃcient estimates.
```{r}
out=glmnet(x,y,alpha=0)
predict (out, type="coefficients", s=bestlam)[1:20,] 
```
None of the coeﬃcients are zero—ridge regression does not perform variable selection.

###The Lasso
Ridge regression with a wise choice of λ can outperform least squares as well as the null model on the Hitters data set. Now ask whether the lasso can yield either a more accurate or a more interpretable model than ridge regression. In order to ﬁt a lasso model, use the glmnet() function; however, this time we use the argument alpha=1. Other than that change,  proceed just as was done ﬁtting a ridge model.
```{r}
lasso.mod=glmnet(x[train,],y[train],alpha=1,lambda=grid)
plot(lasso.mod)
```

See from the coeﬃcient plot that depending on the choice of tuning parameter some of the coeﬃcients will be exactly equal to zero.  Now perform cross-validation and compute the associated test error.
```{r}
set.seed(1)
cv.out=cv.glmnet(x[train,],y[train],alpha=1)
plot(cv.out)
bestlam =cv.out$lambda.min
lasso.pred=predict(lasso.mod ,s=bestlam ,newx=x[test,])
mean((lasso.pred -y.test)^2)
```
This is substantially lower than the test set MSE of the null model and of least squares and very similar to the test MSE of ridge regression with λ chosen by cross-validation. However, the lasso has a substantial advantage over ridge regression in that the resulting coeﬃcient estimates are sparse. 12 of the 19 coeﬃcient estimates are exactly zero. So the lasso model with λ chosen by cross-validation contains only seven variables.
```{r}
out=glmnet(x,y,alpha=1,lambda=grid)
lasso.coef=predict(out, type="coefficients", s=bestlam)[1:20,]
lasso.coef
```
##PCR and PLS Regression

###Principal Components Regression
Principal components regression (PCR) can be performed using the pcr() function which is part of the pls library. Apply PCR to the Hitters data in order to predict Salary. Ensure that the missing values have been removed from the data.
```{r}
library(pls)
set.seed(2)
pcr.fit = pcr(Salary~., data = Hitters, scale = TRUE, valiadation = "CV")
```
The syntax for the pcr() function is similar to that for lm() with a few additional options. Setting scale=TRUE has the eﬀect of standardizing each predictor prior to generating the principal components so that the scale on which each variable is measured will not have an eﬀect. Setting validation="CV" causes pcr() to compute the ten-fold cross-validation error for each possible value of M, the number of principal components used. The resulting ﬁt can be examined using summary().
```{r}
summary(pcr.fit)
```
The CV score is provided for each possible number of components ranging from M = 0 onwards. Note pcr() reports the root mean squared error; in order to obtain the usual MSE, square this quantity. For instance, a root mean squared error of 352.8 corresponds to an MSE of 352.8~2~ = 124,468. One can also plot the cross-validation scores using the validationplot() function. Using val.type="MSEP" will cause the cross-validation MSE to be plotted.
```{r}
validationplot(pcr.fit ,val.type="MSEP")
```

The smallest cross-validation error occurs when M = 16 components are used. This is barely fewer than M = 19 which amounts to simply performing least squares because when all of the components are used in PCR no dimension reduction occurs. However, from the plot see that the cross-validation error is roughly the same when only one component is included in the model. This suggests that a model that uses just a small number of components might suﬃce. 

The summary() function also provides the percentage of variance explained in the predictors and in the response using diﬀerent numbers of components. Think of this as the amount of information about the predictors or the response that is captured using M principal components. For example, setting M = 1 only captures 38.31% of all the variance, or information, in the predictors. In contrast, using M = 6 increases the value to 88.63%. If all M = p = 19 components this would increase to 100%. Perform PCR on the training data and evaluate its test set performance.
```{r}
set.seed(1)
pcr.fit = pcr(Salary~., data = Hitters, subset = train, scale = TRUE, valiadation = "CV")
validationplot(pcr.fit ,val.type="MSEP")
```

Now ﬁnd the lowest cross-validation error occurs when M =7 component are used. Compute the test MSE as follows.
```{r}
pcr.pred=predict(pcr.fit , x[test ,], ncomp =7)
mean((pcr.pred -y.test)^2)
```
This test set MSE is competitive with the results obtained using ridge regression and the lasso. However, as a result of the way PCR is implemented, the ﬁnal model is more diﬃcult to interpret because it does not perform any kind of variable selection or even directly produce coeﬃcient estimates. Finally, ﬁt PCR on the full data set, using M = 7, the number of components identiﬁed by cross-validation.
```{r}
pcr.fit=pcr(y∼x,scale=TRUE,ncomp=7)
summary (pcr.fit)
```

###Partial Least Squares
Implement partial least squares (PLS) using the plsr() function, also in the pls library. The syntax is just like that of the pcr() function.
```{r}
set.seed(1)
pls.fit = plsr(Salary~., data = Hitters, subset = train, scale = TRUE, valiadation = "CV")
summary(pls.fit) 
```
The lowest cross-validation error occurs when only M = 2 partial least squares directions are used. Evaluate the corresponding test set MSE.
```{r}
pls.pred=predict(pls.fit, x[test ,], ncomp =2)
mean((pls.pred-y.test)^2)
```
The test MSE is comparable to, but slightly higher than, the test MSE obtained using ridge regression, the lasso, and PCR. Finally,  perform PLS using the full data set, using M = 2, the number of components identiﬁed by cross-validation.
```{r}
pls.fit = plsr(Salary~., data = Hitters, scale = TRUE, ncomp = 2)
summary(pls.fit) 
```
Notice that the percentage of variance in Salary that the two-component PLS ﬁt explains, 46.40%, is almost as much as that explained using the ﬁnal seven-component model PCR ﬁt, 46.69%. This is because PCR only attempts to maximize the amount of variance explained in the predictors, while PLS searches for directions that explain variance in both the predictors and the response.
