---
title: "ISLR - CH 7 - Beyond Linearity"
output: html_document
---

Non-linear fitting procedures discussed can be easily implemented in R.
```{r}
library(ISLR)
attach(Wage)
```
Fit the model using the following command:
```{r}
fit = lm(wage~poly(age, 4), data = Wage)
coef(summary(fit))
```
This syntax fits a linear model using the lm() function in order to predict wage using a fourth-degree polynomial in age: poly(age, 4). The poly() command allows us to avoid having to write out a long formula with powers of age. The function returns a matrix whose columns are a basis of orthogonal polynomials which  means  each column is a linear combination of the variables age, age^2, age^3 and age^4.

However, we can also use poly() to obtain age, age^2, age^3 and age^4 directly. Do this by using the raw=TRUE argument to the poly() function. This does not affect the model in a meaningful way—though the choice of basis clearly affects the coefficient estimates, it does not affect the fitted values obtained.
```{r}
fit2=lm(wage~poly(age, 4, raw =T), data = Wage)
coef(summary(fit2))
```
There are several other equivalent ways of fitting this model which showcase the flexibility of the formula language in R. For example:
```{r}
fit2a = lm(wage~age + I(age^2) + I(age^3) +I (age^4), data = Wage)
coef(fit2a)
```
This simply creates the polynomial basis functions on the fly taking care to protect terms like age^2 via the wrapper function I() (the ^ symbol has a special meaning in formulas).
```{r}
fit2b = lm(wage~cbind(age, age^2, age^3, age^4), data = Wage)
```
This does the same more compactly using the cbind() function for building a matrix from a collection of vectors; any function call such as cbind() inside a formula also serves as a wrapper.

Create a grid of values for a given age for  predictions and then call the generic predict() function specifying  standard errors.
```{r}
agelims = range(age)
age.grid = seq(from = agelims[1], to = agelims[2])
preds = predict(fit, newdata = list(age = age.grid), se = TRUE)
se.bands = cbind(preds$fit + 2 * preds$se.fit, preds$fit - 2 * preds$se.fit)
```
Plot the data and add the fit from the degree-4 polynomial.
```{r}
par(mfrow = c(1, 2), mar =c (4.5, 4.5, 1, 1), oma = c(0, 0, 4, 0))
plot(age, wage, xlim = agelims, cex = .5, col ="darkgrey")
title("Degree -4 Polynomial", outer = T)
lines(age.grid, preds$fit, lwd = 2, col ="blue")
matlines(age.grid, se.bands, lwd = 1, col ="blue", lty = 3)
```

The mar and oma arguments to par() allow control of the margins of the plot and the title() function creates a figure that spans both subplots.

Whether or not an orthogonal set of basis functions is produced in the poly() function affects the model obtained in a meaningful way. The fitted values obtained in either case are identical:
```{r}
preds2 = predict(fit2, newdata = list(age = age.grid), se = TRUE)
max(abs(preds$fit - preds2$fit))
```
In performing a polynomial regression the degree of the polynomial to use must be determined. One way to do this is by using hypothesis tests. Fit models ranging from linear to a degree-5 polynomial and seek to determine the simplest model which is sufficient to explain the relationship between wage and age. Use the anova() function, which performs an analysis of variance (ANOVA, using an F-test) in order to test the null hypothesis that a model M~1~ is sufficient to explain the data against the alternative hypothesis that a more complex model M~2~ is required. In order to use the anova() function, M~1~ and M~2~ must be nested models: the predictors in M~1~ must be a subset of the predictors in M~2~. In this case, fit five different models and sequentially compare the simpler model to the more complex model.
```{r}
fit.1 = lm(wage~age, data = Wage)
fit.2 = lm(wage~poly(age, 2), data = Wage)
fit.3 = lm(wage~poly(age, 3), data = Wage)
fit.4 = lm(wage~poly(age, 4), data = Wage)
fit.5 = lm(wage~poly(age, 5), data = Wage)
anova(fit.1, fit.2, fit.3, fit.4, fit.5)
```
The p-value comparing the linear Model 1 to the quadratic Model 2 is essentially zero (<10^−15^) indicating that a linear fit is not sufficient. Similarly the p-value comparing the quadratic Model 2 to the cubic Model 3 is very low (0.0017) so the quadratic fit is also insufficient. The p-value comparing the cubic and degree-4 polynomials, Model 3 and Model 4, is approximately 5% while the degree-5 polynomial Model 5 seems unnecessary because its p-value is 0.37. Hence, either a cubic or a quartic polynomial appear to provide a reasonable fit to the data but lower- or higher-order models are not justified.

In this case, instead of using the anova() function, we could have obtained these p-values more succinctly by exploiting the fact that poly() creates orthogonal polynomials.
```{r}
coef(summary(fit.5))
```
Notice the p-values are the same and the square of the t-statistics are equal to the F-statistics from the anova() function; for example:
```{r}
(-11.983)^2
```
However, the ANOVA method works whether or not orthogonal polynomials are used; it also works when other terms in the model as well. For example, we can use anova() to compare these three models:
```{r}
fit.1 = lm(wage~education + age, data = Wage)
fit.2 = lm(wage~education + poly(age, 2), data = Wage)
fit.3 = lm(wage~education + poly(age, 3), data = Wage)
anova(fit.1, fit.2, fit.3)
```
As an alternative to using hypothesis tests and ANOVA, the polynomial degree using cross-validation could be used.

Consider the task of predicting whether an individual earns more than $250,000 per year. Proceed much as before except that first create the appropriate response vector and then apply the glm() function using family="binomial" in order to fit a polynomial logistic regression model.
```{r}
fit = glm(I(wage > 250)~poly(age, 4), data = Wage, family = binomial)
```
The wrapper I() is used to create this binary response variable on the fly. The expression wage > 250 evaluates to a logical variable containing TRUEs and FALSEs which glm() coerces to binary by setting the TRUEs to 1 and the FALSEs to 0.
Make predictions using the predict() function.
```{r}
preds = predict(fit, newdata = list(age = age.grid), se=TRUE)
```
Calculating the confidence intervals is slightly more involved than in the linear regression case. The default prediction type for a glm() model is type="link".

The standard errors given are also of this form. In order to obtain confidence intervals for Pr(Y = 1|X), use the transformation:
Pr(Y = 1|X) =exp(Xβ)/1 + exp(Xβ)
```{r}
pfit = exp(preds$fit )/(1 + exp(preds$fit ))
se.bands.logit = cbind(preds$fit + 2 * preds$se.fit, preds$fit - 2 * preds$se.fit)
se.bands = exp(se.bands.logit)/(1 + exp(se.bands.logit))
```
Note the probabilities could be directly compu ted by selecting the type="response" option in the predict() function.
```{r}
preds = predict(fit, newdata = list(age = age.grid), type ="response", se=TRUE)
```
However, the corresponding confidence intervals would not have been sensible because it produces negative probabilities!

Finally, create the plot:
```{r}
plot(age, I(wage > 250), xlim=agelims, type = "n", ylim = c(0, .2))
points(jitter(age), I((wage > 250) / 5) , cex = .5, pch = "|", col = "darkgrey")
lines(age.grid, pfit, lwd = 2, col = "blue")
matlines(age.grid, se.bands, lwd = 1, col = "blue", lty = 3)
```

The age values have been drawn corresponding to the observations with wage values above 250 as gray marks on the top of the plot, and those with wage values below 250 are shown as gray marks on the bottom of the plot. jitter() function was used to jitter the age values a bit so that observations with the same age value do not cover each other up. This is often called a rug plot.

In order to fit a step function use thecut() function.

```{r}
table(cut(age, 4))
fit = lm(wage~cut(age, 4), data = Wage)
coef(summary(fit))
```
The cut() automatically picked the cutpoints at 33.5, 49, and 64.5 years of age. Cutpoints can be specified directly using the breaks option. The function cut() returns an ordered categorical variable; the lm() function then creates a set of dummy variables for use in the regression. The age < 33.5 category is left out so the intercept coefficient of $94,160 can be interpreted as the average salary for those under 33.5 years of age and the other coefficients can be interpreted as the average additional salary for those in the other age groups.

##Splines

To fit regression splines in R, use the splines library. The bs() function generates the entire matrix of basis functions for splines with the specified set of knots. By default, cubic splines are produced. Fitting wage to age using a regression spline is simple:
```{r}
library(splines)
fit = lm(wage~bs(age, knots = c(25, 40, 60)), data = Wage)
pred = predict(fit, newdata = list(age = age.grid), se = TRUE)
plot(age, wage, col = "gray")
lines(age.grid, pred$fit, lwd = 2)
lines(age.grid, pred$fit + 2 * pred$se, lty = "dashed")
lines(age.grid, pred$fit - 2 * pred$se, lty = "dashed")
```

This produces a spline with six basis functions. (Recall that a cubic spline with three knots has seven degrees of freedom; these degrees of freedom are used up by an intercept, plus six basis functions.) Could also use the df option to produce a spline with knots at uniform quantiles of the data.
```{r}
dim(bs(age, knots = c (25, 40, 60)))
dim(bs(age, df =6 ))
attr(bs(age, df = 6), "knots")
```
In this case R chooses knots at ages 33.8, 42.0, and 51.0, which correspond to the 25th, 50th, and 75th percentiles of age. The function bs() also has a degree argument so we can fit splines of any degree rather than the default degree of 3 (which yields a cubic spline).

In order to instead fit a natural spline use the ns() function. Fit a natural spline with four degrees of freedom.
```{r}
plot(age, wage, col = "gray") #added this to make the RMD process
fit2 = lm(wage~ns(age, df = 4), data = Wage)
pred2 = predict(fit2, newdata = list(age = age.grid), se = TRUE)
lines(age.grid, pred2$fit, col = "red", lwd = 2)
```

As with the bs() function, the knots can be specified directly using the knots option.

To fit a smoothing spline, use the smooth.spline() function.
```{r}
plot(age, wage, xlim = agelims, cex = .5, col = "darkgrey")
title("Smoothing Spline")
fit = smooth.spline(age, wage, df = 16)
fit2 = smooth.spline(age, wage, cv = TRUE)
fit2$df

lines(fit, col = "red", lwd = 2)
lines(fit2, col = "blue", lwd = 2)
legend("topright", legend = c("16 DF", "6.8 DF"), col=c("red", "blue"), lty = 1, lwd = 2, cex = .8)
```

Notice that in the first call to smooth.spline(), df=16 is specified. The function then determines which value of λ leads to 16 degrees of freedom. In the second call to smooth.spline(), we select the smoothness level by crossvalidation; this results in a value of λ that yields 6.8 degrees of freedom.

In order to perform local regression, use the loess() function.
```{r}
plot(age, wage, xlim = agelims, cex = .5, col = "darkgrey")
title("Local Regression")
fit=loess(wage~age, span = .2, data = Wage)
fit2 = loess(wage~age, span = .5, data = Wage)
lines(age.grid, predict(fit, data.frame(age = age.grid)), col ="red", lwd = 2)
lines(age.grid, predict(fit2, data.frame(age = age.grid)), col = "blue", lwd = 2)
legend("topright", legend = c("Span = 0.2", "Span = 0.5"), col=c("red", "blue"), lty = 1, lwd = 2, cex = .8)
```

Performed local linear regression using spans of 0.2 and 0.5: that is, each neighborhood consists of 20% or 50% of the observations. The larger the span, the smoother the fit. The locfit library can also be used for fitting local regression models in R.

###GAMs

Fit a GAM to predict wage using natural spline functions of year and age, treating education as a qualitative predictor. Sincethis is just a big linear regression model using an appropriate choice of basis functions, simply use the lm() function.
```{r}
gam1 = lm(wage~ns(year, 4) + ns(age, 5) + education, data = Wage)
```
Now fit using smoothing splines rather than natural splines. In order to fit more general sorts of GAMs, using smoothing splines or other components that cannot be expressed in terms of basis functions and then fit using least squares regression, we will need to use the gam library in R.

The s() function, which is part of the gam library, is used to indicate using a smoothing spline. Specify the function of year should have 4 degrees of freedom and that the function of age will have 5 degrees of freedom. Since education is qualitative, we leave it as is and it is converted into four dummy variables. We use the gam() function in order to fit a GAM using these components. 
```{r}
library(gam)
gam.m3 = gam(wage~s(year, 4) + s(age, 5) + education, data = Wage)
```
Call the plot() function:
```{r}
par(mfrow = c(1, 3))
plot(gam.m3, se = TRUE, col = "blue")
```

The generic plot() function recognizes gam.m3 is an object of class gam and invokes the appropriate plot.gam() method. Conveniently, even though gam1 is not of class gam but rather of class lm, we can still use plot.gam() on it. 
```{r}
plot.gam(gam1, se = TRUE, col = "red")
```

Notice plot.gam() was used rather than the generic plot() function.

In these plots the function of year looks rather linear. Perform a series of ANOVA tests in order to determine which of these three models is best: a GAM that excludes year (M~1~), a GAM that uses a linear function of year (M~2~), or a GAM that uses a spline function of year (M~3~). 
```{r}
gam.m1 = gam(wage~s(age, 5) + education, data = Wage)
gam.m2 = gam(wage~year + s(age, 5) + education, data = Wage)
gam.m3 = gam(wage~s(year, 4) + s(age, 5) + education, data = Wage)#added this - appears to be missing from book
anova(gam.m1, gam.m2, gam.m3, test = "F")
```
There is compelling evidence that a GAM with a linear function of year is better than a GAM that does not include year at all (p-value=0.00014). However, there is no evidence that a non-linear function of year is needed (p-value=0.349). In other words, based on the results of this ANOVA, M~2~ is preferred.

The summary() function produces a summary of the gam fit.
```{r}
summary(gam.m3)
```
The p-values for year and age correspond to a null hypothesis of a linear relationship versus the alternative of a non-linear relationship. The large p-value for year reinforces our conclusion from the ANOVA test that a linear function is adequate for this term. However, there is very clear evidence that a non-linear term is required for age. Can make predictions from gam objects, just like from lm objects, using the predict() method for the class gam. Make predictions on the training set.
```{r}
preds = predict(gam.m2, newdata = Wage)
```
Can also use local regression fits as building blocks in a GAM, using the lo() function.
```{r}
gam.lo= gam(wage~s(year, df = 4) + lo(age, span = 0.7) + education, data = Wage)
plot.gam(gam.lo, se = TRUE, col = "green")
```

Here local regression was used for the age term with a span of 0.7. Can also use the lo() function to create interactions before calling the gam() function. For example,
```{r}
gam.lo.i = gam(wage~lo(year, age, span = 0.5) + education, data = Wage)
```
fits a two-term model in which the first term is an interaction between year and age, fit by a local regression surface. Plot the resulting two-dimensional surface afterinstalling the akima package.
```{r}
library(akima)
plot(gam.lo.i)
```

In order to fit a logistic regression GAM, use the I() function in constructing the binary response variable and set family=binomial.
```{r}
gam.lr = gam(I(wage > 250)~year +s (age, df = 5) + education, family = binomial, data = Wage)
par(mfrow = c(1,3))
plot(gam.lr, se =TRUE, col = "green")
```

It is easy to see that there are no high earners in the less than HS category:
```{r}
table(education, I(wage > 250))
```

Fit a logistic regression GAM using all but this category. This provides more sensible results.
```{r}
gam.lr.s = gam(I(wage > 250)~year + s(age, df = 5) + education, family = binomial, data = Wage, 
               subset = (education != "1. < HS Grad"))
plot(gam.lr.s, se= TRUE, col = "green")
```