---
title: "ISLR - Ch 5 - Resampling Methods"
output: html_document
---

##Cross-Validation and the Bootstrap

###The Validation Set Approach

Begin by using the sample() function to split the set of observations into two halves by selecting a random subset of 196 observations out of the original 392 observations.
```{r}
library (ISLR)
set.seed (1)
train=sample(392, 196)
```

Use the subset option in lm() to fit a linear regression using only the observations corresponding to the training set.
```{r}
lm.fit = lm(mpg ~ horsepower, data = Auto, subset = train)
```
Use the predict() function to estimate the response for all 392 observations, and we use the mean() function to calculate the MSE of the 196 observations in the validation set. Note that the -train index below selects only the observations that are not in the training set.
```{R}
attach(Auto)
mean((mpg - predict(lm.fit ,Auto))[-train ]^2)
```
The estimated test MSE for the linear regression fit is 26.14. Use the poly() function to estimate the test error for the polynomial and cubic regressions.
```{r}
lm.fit2 = lm(mpg ~ poly(horsepower, 2), data = Auto, subset = train)
mean((mpg - predict(lm.fit2, Auto))[-train ]^2)

lm.fit3 = lm(mpg ~ poly(horsepower, 3), data = Auto,subset = train)
mean((mpg - predict(lm.fit3, Auto))[-train ]^2)
```
The error rates are 19.82 and 19.78. If a different training set is used instead different errors on the validation set will result.
```{r}
set.seed(2)
train = sample(392, 196)
lm.fit = lm(mpg ~ horsepower,subset = train)
mean((mpg -predict (lm.fit ,Auto))[-train ]^2)

lm.fit2 = lm(mpg ~ poly(horsepower, 2), data = Auto, subset = train)
mean((mpg - predict(lm.fit2, Auto))[-train ]^2)

lm.fit3 = lm(mpg ~ poly(horsepower, 3), data = Auto, subset = train)
mean((mpg  - predict(lm.fit3, Auto))[-train ]^2)
```
Using this split of the observations into a training set and a validation set, the validation set error rates for the models with linear, quadratic, and cubic terms are 23.30, 18.90, and 19.26, respectively. These results are consistent with previous findings: a model that predicts mpg using a quadratic function of horsepower performs better than a model that involves only a linear function of horsepower and there is little evidence in favor of a model that uses a cubic function of horsepower. 

###5.3.2 Leave-One-Out Cross-Validation

The LOOCV estimate can be automatically computed for any generalized linear model using the glm() and cv.glm() functions. If glm() is used to fit a model without passing in the family argument then it performs linear regression just like the lm() function. So for instance,
```{r}
glm.fit = glm(mpg ~ horsepower, data = Auto)
coef(glm.fit)
```
and
```{r}
lm.fit = lm(mpg ~ horsepower, data = Auto)
coef(lm.fit)
```
yield identical linear regression models. Using the glm() function to perform linear regression rather than the lm() function is preferred because the latter can be used together with cv.glm(). The cv.glm() function is part of the boot library.
```{r}
library (boot)
glm.fit = glm(mpg ~ horsepower, data = Auto)
cv.err = cv.glm(Auto, glm.fit)
cv.err$delta
```
The cv.glm() function produces a list with several components. The two numbers in the delta vector contain the cross-validation results. In this case the numbers are identical (up to two decimal places). The cross-validation estimate for the test error is approximately 24.23.

Repeat this procedure for increasingly complex polynomial fits. To automate the process, use the for() function to initiate a for loop which iteratively fits polynomial regressions for polynomials of order i = 1 to i = 5, computes the associated cross-validation error and stores it in the ith element of the vector cv.error. Begin by initializing the vector. This command will likely take a couple of minutes to run.
```{r}
cv.error = rep(0,5)
for (i in 1:5){
   glm.fit = glm(mpg ~ poly(horsepower, i), data = Auto)
   cv.error[i] = cv.glm(Auto, glm.fit)$delta[1]}

cv.error
```
There is a sharp drop in the estimated test MSE between the linear and quadratic fits but then no clear improvement from using higher-order polynomials.

###k-Fold Cross-Validation
The cv.glm() function can also be used to implement k-fold CV. Use k = 10, a common choice for k, on the Auto data set. Again set a random seed and initialize a vector in which will store the CV errors corresponding to the polynomial fits of orders one to ten. 
```{r}
set.seed(17)
cv.error.10 = rep(0, 10)
for(i in 1:10) {
  glm.fit = glm(mpg ~ poly(horsepower, i),data = Auto)
  cv.error.10[i] = cv.glm(Auto, glm.fit, K = 10)$delta[1]}

cv.error.10
```
The computation time is much shorter than that of LOOCV. There is little evidence that using cubic or higher-order polynomial terms leads to lower test error than simply using a quadratic fit.

Two numbers associated with delta are essentially the same when LOOCV is performed. Using k-fold CV instead,  the two numbers associated with delta differ slightly. The first is the standard k-fold CV estimate. The second is a bias corrected version. On this data set, the two estimates are very similar to each other.

###The Bootstrap

####Estimating the Accuracy of a Statistic of Interest

One of the great advantages of the bootstrap approach is that it can be applied in almost all situations. No complicated mathematical calculations are required. Performing a bootstrap analysis in R entails only two steps. First, create a function that computes the statistic of interest. Second, use the boot() function which is part of the boot library, to perform the bootstrap by repeatedly sampling observations from the data set with replacement.

To illustrate the use of the bootstrap on this data, create a function, alpha.fn(), which takes as input the (X, Y) data as well as a vector indicating which observations should be used to estimate α. The function then outputs the estimate for α based on the selected observations.
```{r}
alpha.fn = function(data, index){
  X = data$X[index]
  Y = data$Y[index]
  return ((var(Y) - cov(X,Y))/(var(X) + var(Y) - 2*cov(X, Y)))}
```
This function returns an estimate for α. The following command tells R to estimate α using all 100 observations.
```{r}
alpha.fn(Portfolio, 1:100)
```
The next command uses the sample() function to randomly select 100 observations from the range 1 to 100, with replacement. This is equivalent to constructing a new bootstrap data set and recomputing ˆα based on the new data set.
```{r}
set.seed(1)
alpha.fn(Portfolio,sample(100, 100, replace = T))
```
Implement a bootstrap analysis by performing this command many times recording all of the corresponding estimates for α and computing the resulting standard deviation. However, the boot() function automates this approach. Produce R = 1,000 bootstrap estimates for α.
```{r}
boot(Portfolio, alpha.fn, R = 1000)
```
The final output shows that using the original data, ˆα = 0.5758 and the bootstrap estimate for SE(ˆα) is 0.0886.

####Estimating the Accuracy of a Linear Regression Model

The bootstrap approach can be used to assess the variability of the coefficient estimates and predictions from a statistical learning method. Use the bootstrap approach in order to assess the variability of the estimates for β0 and β1, the intercept and slope terms for the linear regression model that uses horsepower to predict mpg in the Auto data set. 

Create a simple function, boot.fn(), which takes in the Auto data set as well as a set of indices for the observations and returns the intercept and slope estimates for the linear regression model. Apply this function to the full set of 392 observations in order to compute the estimates of β0 and β1 on the entire data set using the usual linear regression coefficient estimate formulas. 
```{r}
boot.fn = function(data, index)
     return (coef(lm(mpg∼horsepower ,data=data ,subset =index)))

boot.fn(Auto, 1:392)
```
The boot.fn() function can also be used in order to create bootstrap estimates for the intercept and slope terms by randomly sampling from among the observations with replacement. Two examples:
```{r}
set.seed(1)
boot.fn(Auto, sample(392, 392, replace = T))

boot.fn(Auto, sample(392, 392, replace = T))
```
Use the boot() function to compute the standard errors of 1,000 bootstrap estimates for the intercept and slope terms.
```{r}
boot(Auto, boot.fn, 1000)
```
This indicates that the bootstrap estimate for SE(ˆβ0) is 0.86, and that the bootstrap estimate for SE( ˆ β1) is 0.0074. Standard formulas can be used to compute the standard errors for the regression coefficients in a linear model. These can be obtained using the summary() function.
```{r}
summary(lm(mpg ~ horsepower, data =Auto ))$coef
```
The standard error estimates for ˆβ0 and ˆβ1 are 0.717 for the intercept and 0.0064 for the slope. These are somewhat different from the estimates obtained using the bootstrap. Does this indicate a problem with the bootstrap? In fact, it suggests the opposite. Standard formulas rely on certain assumptions. They depend on the unknown parameter σ2, the noise variance. Then estimate σ2 using the RSS. Although the formula for the standard errors do not rely on the linear model being correct, the estimate for σ2 does. There is a non-linear relationship in the data and the residuals from a linear fit will be inflated, and so will ˆσ2. Also, the standard formulas assume the xi are fixed and all the variability comes from the variation in the errors. The bootstrap approach does not rely on any of these assumptions and so it is likely giving a more accurate estimate of the standard errors of ˆβ0 andˆ β1 than is the summary() function.

Compute the bootstrap standard error estimates and the standard linear regression estimates that result from fitting the quadratic model to the data. Since this model provides a good fit to the data there is now a better correspondence between the bootstrap estimates and the standard estimates of SE(ˆβ0), SE(ˆβ1) and SE(ˆβ2).
```{r}
boot.fn = function(data, index)
  coefficients(lm(mpg~ horsepower + I(horsepower^2), data = data, subset = index))
set.seed(1)
boot(Auto, boot.fn, 1000)
```